\thetitle

### 2023-06-03 lang-design.tex
Fix typo for Prec.

\begin{centered}
\begin{minipage}{\textwidth}
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9.9}\ttfamily]
tm_prect_STLC : \forall (P : tm -> Type), [@\hlcase{tm-partialrec-case-unit}@]option (P tm_unit)[@\tikzmarkend{tm-partialrec-case-unit}@] -> [@\hlcase{tm-partialrec-case-var}@](\forall x, option (P (tm_var x)))[@\tikzmarkend{tm-partialrec-case-var}@] ->
  [@\hlcase{tm-partialrec-case-abs}@](\forall x t, option (P t) -> option (P (tm_abs x t)))[@\tikzmarkend{tm-partialrec-case-abs}@] ->
  [@\hlcase{tm-partialrec-case-app}@](\forall t1, option (P t1) -> \forall t2, option (P t2) -> option (P (tm_app t1 t2)))[@\tikzmarkend{tm-partialrec-case-app}@] -> \forall t, option (P t)
(* For arbitrary [@$P$@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@], and we denote [@"tmPR"@] := [@"tm_prect_STLC"@] [@$P$@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@] *)
(* We will have following equations *)
tm_prect_STLC_unit : tmPR tm_unit      = [@$r_1$@]
tm_prect_STLC_var  : tmPR (tm_var x)   = [@$r_2$@] x
tm_prect_STLC_abs  : tmPR (tm_abs x b) = [@$r_3$@] a b (tmPR b)
tm_prect_STLC_app  : tmPR (tm_app a b) = [@$r_4$@] a (tmPR a) b (tmPR b)
\end{lstlisting}
\end{minipage}
\end{centered}

This corresponds to the dependent eliminator for the non-extensible version of "tm".

@@

  \begin{itemize}
    \item \textlabel{(Detectable Partiality)}{prop:detectable-partiality} for each pair $x : A_i \vdash B_i(x)$,
    \begin{minted}{Coq}
      Axiom detectable:
      forall {T} {a} (f: Bᵢ a -> option T),
        {forall x, f x <> None} + {exists x, f x = None}.
    \end{minted} 
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)), ..,@\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}\YZreply{Is this constraint Detectable Partiality?}@
      D -> option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using the eliminator of\/
  "C" and "prec" of\/ "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}
\begin{proof} 
First, for a fixed \mintinline{Coq}{x : Aᵢ}, given \mintinline{Coq}{fᵢ :  Bᵢ x -> T},
\\ we have \mintinline{Coq}{lift fᵢ : Bᵢ x -> option T}, simply wrapping option around the returning result.

@@

Theorem detect_lifted:
  forall {a} {T : B a -> Type} (f: (forall c: B a, option (T c))),
    { g | f = lift g} + {exists x, f x = None}.
\end{minted}
The idea of "unlift" is that if "f" never returns "None", then it is always a result of a "lift". "detect_lifted" is just another way of saying  \ref{prop:detectable-partiality}.

Then we can construct "inj" and "linv" as follows, using eliminator and partial recursor.
\begin{minted}{Coq}
Definition inj : C -> D :=
  C_rect (fun (_ : C) ⇒ D) 
    ...
    (fun (xᵢ : Aᵢ) (_ : Bᵢ x -> Cᵢ) (Xᵢ : Bᵢ x -> D) ⇒ dᵢ xᵢ Xᵢ)
    ... .

Definition linv : D -> option C :=
  prec C 
    ...
    (fun (aᵢ : Aᵢ) (bᵢ' : Bᵢ aᵢ → option C) ⇒
       match (detectable bᵢ') with
       | left h =>
           let (bᵢ, _) := unlift bᵢ' h in Some (cᵢ aᵢ bᵢ)
       | right _ => None
       end)
    ... .
\end{minted}
Since each ``constructor'' in "C" and "D" are well-correpsonded, when invoking "C_rect" and "prec", each case is handled similarly, and they only vary along the $\{ "A"_i \vdash "B"_i \}_{i}$ types.

@@

\begin{minted}{Coq}
(* inj_compute can be proved thanks to 
    the computational behaviour about C_rect built-in Coq *)
Lemma inj_compute  : forall a b,
  inj (cᵢ a b)  = dᵢ a (fun x => inj (b x)).

(* linv_compute can be proved thanks to 
    the n (propositional) computational axioms about prec *)
Lemma linv_compute : forall a b,
  linv (dᵢ a b) = 
    let b' := fun x => linv (b x) in 
    match detect_lifted b' with
    | inleft (exist _ x _) => Some (cᵢ a x)
    | inright _ => None
    end.
\end{minted}

Finally,  these two equations is enough to prove that "linv" is really a left inverse, via induction
\begin{minted}{Coq}
linv (inj (cᵢ a b)) 
= linv (dᵢ a (fun x ⇒ inj (b x)))         (* by inj_compute *)
= Some (cᵢ a (fun x ⇒ linv (inj (b x))))  (* by linv_compute *)
= Some (cᵢ a (fun x => b x))              (* by I.H. *)
= Some (cᵢ a b)                           (* by eta rule *)
\end{minted}
\end{proof}

@@

We can support partial dependent eliminator as well, but partial dependent eliminator can deduce "prec" and our "prec" is enough to contruct this left inverse for (non-indexed) inductive type. 
\begin{minted}{Coq}
  pdelim : ∀ (P : D -> Set), .., 
  (rᵢ : (∀ (x : Aᵢ) (w : Bᵢ x -> D), (∀ (b : Bᵢ x), option (P (w b))) 
    -> option (P (dᵢ x w)))),  .., ∀ (d : D) -> option (P d)
\end{minted}
To generalize to indexed inductive type, we need to base on indexed W type~\cite{martin1982constructive, morris2009indexed,jashug2017}. We have mechanized the argument for indexed inductive type and dependent eliminator, but for the case we only have one constructor in "C" and "D". This mechanization is a direct generalization of the proof given in

As expected, "tm_prect_STLC" is axiomatized along with
four equalities describing its computational behaviors, one for each constructor.
For instance, the equality for constructor "tm_abs" is as follows:

>>>>

\begin{centered}
\begin{minipage}{\textwidth}
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9.9}\ttfamily]
tm_prect_STLC : \forall (P : tm -> Type), [@\hlcase{tm-partialrec-case-unit}@]option (P tm_unit)[@\tikzmarkend{tm-partialrec-case-unit}@] -> [@\hlcase{tm-partialrec-case-var}@](\forall x, option (P (tm_var x)))[@\tikzmarkend{tm-partialrec-case-var}@] ->
  [@\hlcase{tm-partialrec-case-abs}@](\forall x t, option (P t) -> option (P (tm_abs x t)))[@\tikzmarkend{tm-partialrec-case-abs}@] ->
  [@\hlcase{tm-partialrec-case-app}@](\forall t1, option (P t1) -> \forall t2, option (P t2) -> option (P (tm_app t1 t2)))[@\tikzmarkend{tm-partialrec-case-app}@] -> \forall t, option (P t)
(* For arbitrary [@$P$@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@], and we denote [@"tmPR"@] := [@"tm_prect_STLC"@] [@$P$@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@] *)
(* We will have following equations *)
tm_unit_eq_STLC : tmPR tm_unit      = [@$r_1$@]
tm_var_eq_STLC  : tmPR (tm_var x)   = [@$r_2$@] x
tm_abs_eq_STLC  : tmPR (tm_abs x b) = [@$r_3$@] a b (tmPR b)
tm_app_eq_STLC  : tmPR (tm_app a b) = [@$r_4$@] a (tmPR a) b (tmPR b)
\end{lstlisting}
\end{minipage}
\end{centered}

These four equations uniquely describe the "tm_prect_STLC" on the four constructors and thus "tm_prect_STLC" cannot be a trivial function. This partial recursor also corresponds to the dependent eliminator for the non-extensible version of "tm".

@@

  \begin{itemize}
    \item \textlabel{(Detectable Partiality)}{prop:detectable-partiality} for each pair $x : A_i \vdash B_i(x)$,
    \begin{minted}{Coq}
      Axiom detectable:
      forall {T} {a} (f: Bᵢ a → option T),
        {forall x, f x <> None} + {exists x, f x = None}.
    \end{minted} 
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x → C) → C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x → D) → D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x → option R) → option R)), ..,@\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}\YZreply{Is this constraint Detectable Partiality?}@
      D → option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x ⇒ prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C → D} and 
  \mintinline{Coq}{linv : D → option C} defined using the eliminator of\/
  "C" and "prec" of\/ "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}
\begin{proof} 
First, for a fixed \mintinline{Coq}{x : Aᵢ}, given \mintinline{Coq}{fᵢ :  Bᵢ x → T},
\\ we have \mintinline{Coq}{lift fᵢ : Bᵢ x → option T}, simply wrapping option around the returning result.

@@

Theorem detect_lifted:
  forall {a} {T : Set} (f: (forall c: B a, option T)),
    { g | f = lift g} + {exists x, f x = None}.
\end{minted}
The idea of "unlift" is that if "f" never returns "None", then it is always a result of a "lift". "detect_lifted" is just another way of saying  \ref{prop:detectable-partiality}.

Then we can construct "inj" and "linv" as follows, using eliminator and partial recursor.
\begin{minted}{Coq}
Definition inj : C → D :=
  C_rect (fun (_ : C) ⇒ D) 
    ...
    (fun (xᵢ : Aᵢ) (_ : Bᵢ x → Cᵢ) (Xᵢ : Bᵢ x → D) ⇒ dᵢ xᵢ Xᵢ)
    ... .

Definition linv : D → option C :=
  prec C 
    ...
    (fun (aᵢ : Aᵢ) (bᵢ' : Bᵢ aᵢ → option C) ⇒
       match (detectable bᵢ') with
       | left h ⇒
           let (bᵢ, _) := unlift bᵢ' h in Some (cᵢ aᵢ bᵢ)
       | right _ ⇒ None
       end)
    ... .
\end{minted}
Since each ``constructor'' in "C" and "D" are well-correpsonded, when invoking "C_rect" and "prec", each case is handled similarly, and they only vary along the $\{ "A"_i \vdash "B"_i \}_{i}$ types.

@@

\begin{minted}{Coq}
(* inj_compute can be proved thanks to 
    the computational behaviour about C_rect built-in Coq *)
Lemma inj_compute  : forall a b,
  inj (cᵢ a b)  = dᵢ a (fun x ⇒ inj (b x)).

(* linv_compute can be proved thanks to 
    the n (propositional) computational axioms about prec *)
Lemma linv_compute : forall a b,
  linv (dᵢ a b) = 
    let b' := fun x ⇒ linv (b x) in 
    match detect_lifted b' with
    | inleft (exist _ x _) ⇒ Some (cᵢ a x)
    | inright _ ⇒ None
    end.
\end{minted}

Finally,  these two equations is enough to prove that "linv" is really a left inverse, via induction
\begin{minted}{Coq}
linv (inj (cᵢ a b)) 
= linv (dᵢ a (fun x ⇒ inj (b x)))         (* by inj_compute *)
= Some (cᵢ a (fun x ⇒ linv (inj (b x))))  (* by linv_compute *)
= Some (cᵢ a (fun x ⇒ b x))              (* by I.H. *)
= Some (cᵢ a b)                           (* by eta rule *)
\end{minted}
\end{proof}

@@

We can support partial dependent eliminator as well, but partial dependent eliminator can deduce "prec" and our "prec" is enough to contruct this left inverse for (non-indexed) inductive type. 
\begin{minted}{Coq}
  pdelim : ∀ (P : D → Set), .., 
  (rᵢ : (∀ (x : Aᵢ) (w : Bᵢ x → D), (∀ (b : Bᵢ x), option (P (w b))) 
    → option (P (dᵢ x w)))),  .., ∀ (d : D) → option (P d)
\end{minted}
To generalize to indexed inductive type, we need to base on indexed W type~\cite{martin1982constructive, morris2009indexed,jashug2017}. We have mechanized the argument for indexed W type and dependent eliminator, but for the case with only one constructor in "C" and "D". This mechanization\footnote{\url{https://gist.github.com/DKXXXL/9b4e48350b966fa3739cfb71f1a9a852}} is a direct generalization of the proof for \cref{thm:prec-complete}.

As mentioned, "tm_prect_STLC" is axiomatized along with
four equalities describing its computational behaviors, one for each constructor.
After unfold the definition of "tmPR", the equality for constructor "tm_abs" is as follows:
